Formation of porous crystals via viscoelastic phase separation

### 2017-01-26 nature/RebuttalLetter.tex

In the new version of the manuscript we have tried to incorporate all the Referee's previous suggestions. 
Following the suggestion, we have explicitly mentioned the work of Poon and coworkers in connection to the Bergeron and Ostwald ripening processes.  
We have also added the two recommended references by Sanz et al. to the new version of the manuscript. 
On the doubt of the referee concerning the novelty of our work, we hope that our new analysis explained above provides a 
new physical insight on the physical condition required for the formation of porous crystal gels besides its new kinetic route. 
We also simplified the title to avoid misunderstanding of our message.

Regarding our experimental protocol, it is indeed reminiscent of salt injection, but differs in two important ways. (i) Rather than an injection salts are introduced through a membrane. This dialysis protocol allows salt introduction without external flow. (ii) We are in a regime where, upon salt introduction, charges are screened throughout the sample almost instantaneously, that is to say in a few Brownian times. This is due to the short length the ions have to diffuse from the reservoir to the top of the sample and to the large diffusion constant ratio between the ions and our large colloidal particles.
We believe this to be the first time this cell design has been applied for confocal microscopy observation at a single-particle level and used to study the formation of porous crystals in gels, which is the key to access the early stage of viscoelastic phase separation at a single particle resolution with full particle tracking. 
For example, the new analysis shown in new Fig.~2 is only possible with this new experimental protocol and full particle tracking.  
The absence of any harmful flow upon the initiation of phase separation and any drift can be clearly seen in Supplementary Movie 1.  
Finally, the last sentence of the abstract has been modified, following the kind suggestion of the Referee.

@@

\noindent
\begin{Large}
{\bf Response to Referee \#3}
\end{Large}

\vskip 0.5cm

We would like to thank the Referee for his/her comments. In the following we respond to the new points raised.

\begin{referee}
\noindent
1) I was aware of the different frame rates on the two movies but I agree that the time stamps are helpful. Figure S5 addresses my concern about reproducibility at the single state point, and the different geometry itself is helpful too. I think the authors should be up front in the manuscript that the mechanism discussed is found in a very specific window (from the figures and SI, I think it is $\phi\sim 0.33$ and $c_p~0.39$ (plus minus) 0.01 mg/g) by moving this point into the abstract.
\end{referee}

\textcolor{blue}{We are glad to hear that the reviewer found that the time stamps are helpful and also Fig. S5 addresses his/her concern.}     
Following the Referee's recommendation, we have put the following sentence in the abstract: ``We clarify the physical requirements necessary for this to happen.'' Although this does not explicitly specify the points we observed crystal gels, our new analysis (see below) 
has revealed the physical conditions required for observing crystal-gel formation. We have also discussed this issue in our revised manuscript with the new analysis shown in a new Fig. 2. We stress that the conditions for crystal-gel formation are not stringent and well within the experimental accessible states, as we previously demonstrated in our reproducibility tests inside a capillary. Our new results show that, at low polymer-concentrations, the network undergoes a structural reorganisation, which results in more compact structures, that are a necessary condition for the appearance of the first crystalline nuclei. To our knowledge, this is the first experimental demonstration of this process: it is also not easy to be captured by computer simulations, as they often neglect hydrodynamic interactions which play a fundamental role in this kinetic process, and no previous experiment has had the capability of tracking particles during the early stages of spinodal decomposition. The absence of any harmful flow upon the initiation of phase separation and any drift can be clearly seen in Supplementary Movie 1.

\begin{referee}
2) The addition of the SI movie 3 and the discussion of Figure 2b addresses this concern for me. It's the percolation that is critical to the authors claim of a crystal gel - the percolated structure needs to persist over long times, otherwise, for example, the gel wouldn't have a finite low frequency elastic modulus. I think SI move 3 does show this.
\end{referee}

\textcolor{blue}{We are glad to hear that the reviewer thought that SI movie 3 and the discussion of Fig. 2b have addressed his/her concern and supported our claim of a crystal gel.}

\begin{referee}
3-1) The data quality of Figure 1c, Figure 1d appear insufficient to classify the $c_p$ = 0.38 mg/mg state point as different from the others. The authors should add replications, errors, and significance testing if they wish to distinguish the dimensionality (Fig 1c) and maximum volume fraction (Fig 1d) of the 0.38 mg/ml state point from the others.

@@

I continue to think that the evidence for different behavior in the $c_p$ = 0.38 mg/g state point from Figure 1 c,d is weak. Supplemental Figure 3 is more convincing that the final state points are different. 
\end{referee}

We fully agree that the evidence from Fig.~1 c,d was not conclusive, but the goal of those figures was not to provide conclusive evidence, but to provide a comprehensive study of the early-stages of spinodal decomposition. In this context the Fig.~1 c,d only provide hints that the $c_p$ = 0.38 mg/g state point behaves differently from other state points. Direct evidence comes from the use of bond-orientational order parameters, presented in Fig. 5 and 6. 
Following the Referee's kind suggestion, we have now moved previous Fig. S3 to a new Fig. 3b. 
We hope that this will make the difference of the $c_p$ = 0.38 mg/mg state point from the others clearer.

\begin{referee}
3-2) (Yet, in the new section, the authors write: "This result clearly indicates that a single state point follows a different arrest mechanism, in which phase separation is arrested by crystallization and not by glassiness." Is that really true? It seems overstated. Don't Figure 1 c,d and SF3 just establish that the end point is different? They don't say anything about the mechanism, which would require kinetic analysis, as comes later. And it doesn't say anything about arrest, which would require dynamics.)
\end{referee}

@@

In the new version of the manuscript we have removed that sentence, and replaced it with:

\emph{
These structural analysis provide some evidence that the state point $\phi\approx 0.33$ and $c_p=0.38\,$mg/g could be following a different arrest mechanism, in which phase separation is arrested by crystallization and not by glassiness. In the reminder of this manuscript we investigate this state more closely, to confirm these early results.}

We also agree that our previous manuscript didn't say much about the actual kinetic mechanism by which crystal-gels form. This point was also raised by Referee \#1 in noting that we didn't provide any additional insight in the phase separation process. In order to fully address this issue, we have implemented a kinetic analysis of the trajectories at all state points, computing the probability of bond-breaking between consecutive frames. This new kinetic analysis reveals that at low-polymer concentrations a new mechanism (stress-driven ageing) is taking place, which has the effect of increasing the average coordination of the colloidal particles in the liquid phase.
We believe this to be a novel observation of this phenomenon. Most past simulation works have neglected, or not fully taken into account, the role of hydrodynamic interactions (HI) in the phase separation process. As was shows for example in Ref.~\cite{furukawa2010key}, the phase separation process without HI produces rather compact structures, which would then easily crystallize to form compact nuclei. This is not what happens if one take HI into consideration. Due to the incompressibility condition, the liquid network has a lower-connectivity, and the viscoelastic phase separation results in the accumulation of mechanical stress on these liquid bonds. Two things can happen depending on the polymer concentration. At high-polymer concentration, the bonds are not-easily broken, and the network which is first formed cannot undergo structural reorganisation. This inevitably leads to gelation. On the other side, if the polymer concentration is low enough, the liquid network can reorganise with a rather sharp 
transition to a more compact network. This condition is necessary for the appearance of the first crystalline nuclei. In our new manuscript, we are able to follow the kinetics of these processes by looking at the time evolution of the bond-breaking probability. The different network states are in fact characterised by different bond-breaking probabilities and by different coordination numbers. In Fig.~2a,b we show these quantities for high-polymer concentration samples, where spinodal decomposition is the only process that occurs. At low-polymer concentrations (Fig.~2c,d) instead, the network undergoes a sharp transition to a state with lower bond-breaking probability and higher connectivity. This transition is the stress-driven aging process, and increases the connectivity of the network. Despite being theoretically predicted, Ref.~\cite{tanaka2000,tanaka2007spontaneous}, this transition was not directly observed before in experiments. 
We stress that for the experimental observation we need to follow particles at a single-particle level from the beginning of phase separation. This was made possible by a new experimental protocol of salt injection combined with confocal microscopy observation.

\begin{referee}
4) Figure S4 is critical to the interpretation, including information about all the transition pathways.
\end{referee}

\textcolor{blue}{We are glad to know that the reviewer thought that Fig. S4 is useful for supporting the interpretation.}

\begin{referee}
 5) The revised Figure 4, along with Figure S4, are very helpful to resolving the relative contributions of the three crystalline pathways. In Figure 4b,c, I still think that surface particles should somehow be distinguished - they represent a distinct state and a kind of consistency check, because the different crystallization pathways should all be mediated by this state. Without including the surface states in the primary figure, I don't see how the reader can inspect Figure S4, and get to the 54/34/11 abundances for direct, Bergeron and Ostwald processes. These abundances are the crucial claim of the paper.
 \end{referee}

 \textcolor{blue}{We are glad to hear that the revised Fig. 4, along with Fig. S4, are very useful to resolving the relative contributions of the three crystalline pathways. Following the kind suggestion of the reviewer, we have now included the fraction of surface particles in Fig. 6b and added the related discussion.}

\begin{referee}
In summary the confirmatory experiments (Fig S5 and Movie S3), the additional analysis of the crystalline state (Fig 3S) and the full transition analysis (rev Fig 4 and F S4) offer significant new support of the authors' claims; the authors have something new relative to prior literature. At this point, I think that the authors should: (i) better acknowledge that what is observed here is found in a quite specific region of the state diagram ($phi\sim 0.33$ and $c_p\sim 0.39$ {plus minus} 0.01 mg/g); (ii) better connect Figure 4 and Figure S4 by accounting for surface states in the former figure.

@@

\end{referee}

We are glad to hear that the Referee thought that the previous revisions offered significant new support of the authors' claims. 
On the comment (i), as we explained above, we have clarified the physical conditions required for crystal-gel formation to take place. We have also discussed how we can widen the region where this is to be observed (see page 14). 
(ii)  \textcolor{blue}{As mentioned above,} following the suggestion, in the new version of the manuscript we follow the Referee's recommendation and include in Fig. 6b the fraction of surface particles, and update the description accordingly.

### 2017-01-26 nature/crystal_gel_r.tex

\begin{abstract}
{\bf
Viscoelastic phase separation of colloidal suspensions can be interrupted to form gels either by glass transition or by crystallization. With a new confocal microscopy protocol, \textcolor{blue}{we follow the entire kinetics of phase separation, from the homogeneous phase to the formation of different arrested states. This allows us to unveil a novel crystallization pathway to sponge-like porous crystal structures in a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation. We show that crystallization is preceded by a structural reorganization of the liquid phase, called stress-driven ageing. Nucleation starts inside this liquid network, but crystals grow past it by direct condensation of the gas phase on their surface, i.e. de-sublimation, driving evaporation of nearby liquid. This produces a network structure different from the original phase separation pattern.} We argue that similar crystal-gel states can be formed as the result of the gas-liquid phase separation of any crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification. This mechanism will provide a novel pathway to form nano-porous crystals of metals and semiconductors without dealloying, which may be important for catalytic, optical, sensing, and filtration applications. 
}
\end{abstract}

>>>>

\begin{abstract}
{\bf
Viscoelastic phase separation of colloidal suspensions can be interrupted to form gels either by glass transition or by crystallization. With a new confocal microscopy protocol, \textcolor{blue}{we follow the entire kinetics of phase separation, from the homogeneous phase to the formation of different arrested states. This allows us to unveil a novel crystallization pathway to sponge-like porous crystal structures in a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation. We show that crystallization is preceded by a structural reorganization of the liquid phase, called stress-driven ageing. Nucleation starts inside this liquid network, but crystals grow past it by direct condensation of the gas phase on their surface, i.e. de-sublimation, driving evaporation of nearby liquid. This produces a network structure different from the original phase separation pattern. We clarify the physical requirements necessary for this to happen.} We argue that similar crystal-gel states can be formed as the result of the gas-liquid phase separation of any crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification. This mechanism will provide a novel pathway to form nano-porous crystals of metals and semiconductors without dealloying, which may be important for catalytic, optical, sensing, and filtration applications. 
}
\end{abstract}
